---
title: "Input & Output"
author: "JHuband"
date: "5/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Input/Output

Programs would not be very useful if they could not share their results with the user or if they did not allow users to provide different inputs.

#### Output

You may have noticed in the previous examples that we used `print()` to display the contents of our variables.  The `print()` statement also can be used to display a string:
```{r}
print("Hello everybody.  I hope you are doing well today!")
```

One drawback of R is that the print statement can only work with one object (e.g., a variable or a string).  If you want to print more than one piece of information, you will need to use `paste()` or `paste0()` to combine the information before printing.

```{r}
len <- 5
wid <- 4
area <- len * wid

print(paste("The area of the rectangle is", area))
```

Another way to have the program print results is to use the `cat()` command.  The `cat()` command automatically converts the objects to characters and concatenates them into a single string.

```{r}
len <- 5
wid <- 4
area <- len * wid

cat("The area of the rectangle is", area)
```

There is one more (somewhat fancy) way to print results.  That is with the `sprintf()` command.  The `sprintf` command allows you to create a sentence with placeholders for the variables.

```{r}
len <- 5
wid <- 4
area <- len * wid

sprintf("The area of the rectangle with length, %d, and width, %d, is %d.", len, wid, area)
```

The `%d` is a generic placeholder for values that are integers only (i.e., no decimal places).  If you have a number with a decimal place, you would `%f` as the placeholder.  For strings, you would use `%s` as the placeholder.

```{r}
Name <- "Louie LeChat"
weight <- 9.65
age <- 8

sprintf("Hello, my name is %s.  I am %d years old and weigh %f pounds", Name, age, weight)
```

Later, we will see how to format the output with sprintf.

#### Reading a Text File

Before you can read a text file into your R program, you must know what format the file is in.  Is it a plain text file or is it in a special format like XML or JSON?  Is the text file organized by lines of text or is it in a table format?  You should know the format of the file before you try reading it into your program.

For now, we are going to look at reading in plain text files.

##### Reading an unformatted text file
I have a file, _austen_emma_excerpt.txt_ that contains the first paragraph of the book _Emma_ by Jane Austen.  We can load the text into R's memory by using the _readLines_ command:

```{r}
raw_text <- readLines("austen_emma_excerpt.txt")
print(raw_text)

```

The `readLines` command will read in the text as separate lines and saves the lines as a list.  

If we prefer to have the text as a single string, we will need to use the `paste` command to concatentate the lines.

```{r}
raw_text <- paste(readLines("austen_emma_excerpt.txt"), collapse="\n")
print(raw_text)
```

The `\n` in the paste command is a formatting character that indicates where new lines will start.
If you want to see the text without the `\n` character, you can use the `cat` command:
```{r}
cat(raw_text)
```




<font color=blue>---------------------------------------------------------------</font>

### Activity:  Reading different formats

Make sure that you have downloaded the file "Books.csv". This file a comma-delimited table with columns for title, author, genre, etc.  You can read the contents of that file with a different read command, specifically:
```
raw_table <- read.csv("Books.csv")

```
Write a program that will read the file "Books.csv" and print the contents to the screen.

<font color=blue>---------------------------------------------------------------</font>

There are other commands for reading formats, like XML, JSON, PDF.  Before we can explore these commands, we need to learn a little bit more about functions and packages in R.
